Towards Artificial Intelligence for Metaphysics: \\ the Case of the Inconsistency in G\"odel's Ontological Argument

### 2016-01-07 Papers/InProgress/IJCAI/paper.tex
Improve introduction

\section{Introduction}
Without exaggeration Kurt G\"{o}del's ontological
argument for the existence of God \cite{GoedelNotes,ScottNotes} is
amongst the most discussed formal proofs in modern literature. A rich
body of publications -- including very recent ones -- present,
discuss, assess, criticise, modify and improve G\"{o}del's original
work (see e.g.~Sobel~\shortcite{Sobel} and Oppy~\shortcite{sep-ontological-arguments} and the
references therein).  In philosophy lectures at universities the
argument is regularly presented as a masterpiece argument in
metaphysics. Since summer 2013, when Benzm\"uller and Woltzenlogel-Paleo~\shortcite{J30,C40} first
reported on their successful initial computational
analysis of the G\"odel's respectively Scott's variants of the proof,
their contribution received a media repercussion on a global scale\footnote{A
  small collection of these reports is available at {\scriptsize
    \url{https://github.com/FormalTheology/GoedelGod/blob/master/Press/LinksToNews.md}}},
and numerous bloggers commented on the proof
\cite{fuhrmann15:_blogg_goedel}.

The in-depth computational analysis presented here substantially
extends previous computer-assisted studies on G\"odel's ontological
argument. Similar to the related work \cite{J30,C40} the analysis has
been conducted with automated theorem provers for classical
higher-order logic (HOL) even though G\"odel's proof is actually
formulated in modal higher-order logic. To bridge between the two
logic formalisms we utilise and further improve the logic embeddings
approach \cite{J23,C40} as also employed in preceding related work.

The novel contributions of this paper include (but are not limited to):

\begin{itemize}
\item A detailed analysis of the inconsistency of G\"{o}del's
  \shortcite{GoedelNotes} original version of the proof is presented
  for higher-order modal logics KB and K (with possibilist and
  actualist first-order quantifiers). The extraction, reconstruction
  and verification of an informal, human intuitive argument has been
  an open problem since the first detection of this inconsistency by
  Benzmüller and Woltzenlogel-Paleo \shortcite{C40}. The experiments
  reported here led to the discovery of a surprisingly accessible
  understanding of this inconsistency which is philosophically
  profound and which has not been presented yet in the respective
  literature. The detection of this inconsistency in combination with
  the work reported here thus demonstrates that higher-order automated
  theorem provers (and thus AI systems) are capable of contributing
  new and relevant knowledge in philosophy. 
\item Another open question has been whether the G\"odel's
  respectively Scott's main theorem T3 \textit{``Necessarily, there
    exists God''} can eventually be proved fully automatically
  directly from the axioms, that is, without the need for intermediate
  argumentation steps (lemmata). Benzm\"uller and Woltzenlogel-Paleo
  \shortcite{C40} still failed in achieving this. We here report
  success for Scott's (consistent) variant of the proof in
  higher-order modal logic S5. The fully automated proof has been
  contributed by the theorem prover \textsc{Leo-II}~\cite{C26} and
  subsequently verified in the proof assistant
  Isabelle/HOL~\cite{NPW02}. On a technical level we show that the
  novel embedding of S5 is more effective than the previous one.
\item A third open issue has been whether the modal
  collapse \cite{Sobel}, which is one of the most strongly criticised
  `side-effects' of G\"odel's and Scott's variants of the proof, can
  eventually be directly verified in Isabelle/HOL. Indeed, in previous work
  the modal collapse has been confirmed by the provers
  \textsc{Satallax} \cite{Satallax} and \textsc{Leo-II}, but a
  direct verification in the highly trusted Isabelle/HOL still failed.
\item On the technical side, we present a significantly improved
  syntax representation of G\"odel's resp.~Scott's proof in
  Isabelle/HOL. In fact, we demonstrate that a nearly perfect match
  between the original pen and paper presentations and our encoding in
  Isabelle/HOl is feasible. This is clearly an important prerequisite
  for promoting the here employed theorem proving technology to
  non-expert philosophers.
\item On the other hand, we will
  also point to several technical challenges which, at least for the
  moment, still restrain non-expert users of these systems from
  independently conducting similar experiments in computational
  metaphysics as reported here and in previous work. 
% Technical issues detected within Isabelle-Sledgehammer; room for improvements
\end{itemize}

@@

First successful applications of theorem proving technology in
metaphysics were reported by Oppenheimer and
Zalta~\shortcite{oppenheimera11}. They employed the first-order
\textsc{Prover9} \cite{Prover9} in their experiments. It was also
Zalta and Oppenheimer who first proposed the term
\textit{Computational Metaphysics} for this new research area. Further
related work includes Rushby's~\shortcite{rushby13} work with the
higher-order proof assistant \textsc{PVS} \cite{PVS}. Common to both
works is a significant amount of proof-hand-coding work as well as their
focus on St. Anselm's~\shortcite{Anselm} simpler, non-modal historical
version of the ontological argument.

\ldots. Different to what is presented here, none of these previous
formalises and automates variants of HOML, which is however crucial
for G\"{o}del's ontological argument.

>>>>

\section{Introduction}
Without exaggeration Kurt G\"{o}del's ontological
argument for the existence of God \cite{GoedelNotes,ScottNotes} is
amongst the most discussed formal proofs in modern literature. A rich
body of publications -- including very recent ones -- present,
discuss, assess, criticise, modify and improve G\"{o}del's original
work (see e.g.~Sobel~\shortcite{Sobel} and Oppy~\shortcite{sep-ontological-arguments} and the
references therein).  In philosophy lectures at universities the
argument is regularly presented as a masterpiece argument in
metaphysics. Since 2013, when Benzm\"uller and Woltzenlogel-Paleo~\shortcite{J30,C40} first
reported their successful initial computer-assisted
analysis of G\"odel's proof and Scott's variant,
their work has received a media repercussion on a global scale\footnote{A
  small collection of news articles is available at {\scriptsize
    \url{https://github.com/FormalTheology/GoedelGod/blob/master/Press/LinksToNews.md}}},
and numerous bloggers commented on the proof
\cite{fuhrmann15:_blogg_goedel}.

The in-depth computational analysis presented here substantially
extends previous computer-assisted studies of G\"odel's ontological
argument. Similar to the related work \cite{J30,C40} the analysis has
been conducted with automated theorem provers for classical
higher-order logic (HOL) even though G\"odel's proof is actually
formulated in \emph{modal} higher-order logic. To bridge between the two
logics we utilise and further improve the logic embedding
approach \cite{J23,C40}, which has already been employed succesfully in preceding related work.

The novel contributions reported in this paper include the following:

\begin{itemize}
\item A detailed analysis of the inconsistency of G\"{o}del's
  \shortcite{GoedelNotes} original version of the proof is presented
  for higher-order modal logics KB and K (with possibilist and
  actualist first-order quantifiers). The extraction, reconstruction
  and verification of an informal, human intuitive argument has been
  an open problem since the first detection of this inconsistency by
  Benzmüller and Woltzenlogel-Paleo \shortcite{C40} with the LEO-II prover. 
  The experiments reported here led to the discovery of a surprisingly accessible
  understanding of this inconsistency which is philosophically
  profound and which has not been presented yet in the
  literature. The detection of this inconsistency in combination with
  the work reported here thus demonstrates that artificial intelligence systems -- 
  particularly higher-order automated
  theorem provers -- are capable of assisting in the discovery and ellucidation of
  \emph{new} and philosophically relevant knowledge. 
\item Another open question has been whether the final theorem T3 \textit{``Necessarily, there
    exists God''} could eventually be proved fully automatically
  directly from the axioms alone. In \shortcite{C40}, the provers had only succeeded in proving 
  T3 from intermediate argumentation steps (lemmata). Here we report that, by using a more efficient new
  embedding for the higher-order modal logic S5, T3 can be directly proven from Scott's version of the axioms, which are consistent. The fully automatic proof has been
  generated by the theorem prover \textsc{Leo-II}~\cite{C26} and
  subsequently verified in the proof assistant
  Isabelle/HOL~\cite{NPW02}.
\item A third open challenge the verification in Isabelle/HOL of the modal
  collapse \cite{Sobel}, which is one of the most strongly criticised
  `side-effects' of G\"odel's and Scott's variants of the proof. In previous work
  the modal collapse has been derived by the provers
  \textsc{Satallax} \cite{Satallax} and \textsc{Leo-II}, but a fully automatic
  verification in the highly trusted Isabelle/HOL still failed. 
  Now, with the more efficient embedding for S5, this verification has been done. 
\item On the technical side, we present a significantly improved
  syntax representation of G\"odel's resp.~Scott's proof in
  Isabelle/HOL. In fact, we demonstrate that a nearly perfect match
  between the original pen and paper presentations and our encoding in
  Isabelle/HOl is feasible. This is clearly an important prerequisite
  for promoting the theorem proving technology employed here to a wide community of
  philosophers, who are not necessarily experts in automated reasoning or higher-order logic.
\item On the other hand, we
  also point to several technical challenges which, at least for the
  moment, still restrain non-expert users of these systems from
  independently conducting similar experiments in computational
  metaphysics as reported here and in previous work. 
% Technical issues detected within Isabelle-Sledgehammer; room for improvements
\end{itemize}

@@

First successful applications of theorem proving technology in
metaphysics were reported by Oppenheimer and
Zalta~\shortcite{oppenheimera11}, who coined the term \textit{Computational Metaphysics} for this new research area and employed the first-order
\textsc{Prover9} \cite{Prover9} in their experiments. Later on, Rushby~\shortcite{rushby13} used the proof assistant \textsc{PVS} \cite{PVS}. Common to both
works is a significant amount of proof-hand-coding work as well as their
focus on a non-modal formalization of St. Anselm's~\shortcite{Anselm} simpler 
and older ontological argument. 
None of these previous works formalises and automates variants of \emph{higher-order} and \emph{modal} logics, which are, however, crucial
for G\"{o}del's more complex ontological argument.
x
